Gains and losses in rheumatoid arthritis data from inferCNV

### 2024-07-09 infercnv/reports/gains_and_losses.Rmd
fix annotations col

---
title: "Gains and losses in rheumatoid arthritis data from inferCNV"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: true
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
  sce: NULL
  cell_group_name: NULL
  infercnv_dir: NULL
  cache_dir: NULL
  rerun: true
  max_prop_for_wt: 0.5
  min_prop_for_cnv: 0.8
---

```{r setup, include = F, message = F, warning = F, class.source = 'fold-hide'}
# wd <- ifelse(Sys.info()['nodename'] == "mib119104s", '~/Volumes/', '') |> paste0('/lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/') ; setwd(wd)
# rmarkdown::render('reports/infercnv.Rmd', output_file = 'driver_coverage.html', output_dir = 'reports/', params = list(rerun = T))

@@

# chunk options
knitr::opts_knit$set(root.dir = '../')
knitr::opts_chunk$set(
  warning = FALSE, 
  dpi = 300, 
  message = FALSE,
  cache.path = params$cache_dir) 
```

@@

```{r load_sce, results = 'asis', class.source = 'fold-show'}
# load gene coords
coords <-
  readr::read_tsv('data/hg38_gencode_v27.txt', col_names = c('name', 'chr', 'start', 'stop'))

# load CTAP and metadata
ctap <-
  readr::read_tsv('data/Zhang2023/syn26710600/CTAP_donor_mapping.tsv') %>%
  dplyr::select(sample = donor, CTAP)
md <- 
  readr::read_tsv('data/Zhang2023/syn26710600/SYNAPSE_METADATA_MANIFEST.tsv') %>%
  dplyr::distinct()

@@

  # get cnv metadata
  cnvs <- 
    list.files(
      params$infercnv_dir,
      pattern = 'map_metadata_from_infercnv.txt',
      recursive = T) %>%
    purrr::set_names(., dirname(.)) %>%
    purrr::map(function(file) {
      file.path(params$infercnv_dir, file) %>% 
        read.table(sep = '\t')
      }) %>%
    dplyr::bind_rows() %>%
    dplyr::mutate(infercnv = T)

  # left join cnvs to coldata, reorder, restore rownames
  col_data <- merge(sce@colData, cnvs, by = 'row.names', all.x = T, sort = F)
  rownames(col_data) <- col_data[, 1]
  col_data <- 
    col_data[colnames(sce), -1] %>%
    DataFrame()

@@

}, file = 'sce.rds', rerun = params$rerun)

@@

```{r echo = F, results = 'asis'}
tibble::tribble(
  ~name, ~`type - value`,
  '`has_*`', 'binary - whether any CNV / loss / duplication is found within the given chr',
  '`proportion_*`', 'proportion - proportion of genes that are part of any CNV / loss / duplication within the given chr',
  '`proportion_scaled_*`', 'proportion - same as proportion_*, but taking into account whether genes that are duplicated / lost are a single copy or two copies, as a weight',
  '`top_loss_*`', 'the top CNVs that affect the most genes'
) %>%
  knitr::kable()
```

@@

  # get HMM predictions
  cnv_regions <- 
    list.files(
      params$infercnv_dir,
      pattern = 'HMM_CNV_predictions.HMMi6.leiden.hmm_mode-subclusters.Pnorm_0.5.pred_cnv_regions.dat',
      recursive = T) %>%
    purrr::set_names(., dirname(.)) %>%
    purrr::map(function(file) {
      file.path(params$infercnv_dir, file) %>% 
        readr::read_tsv(show_col_types = F)
      }) %>%
    dplyr::bind_rows(.id = 'sample') %>%
    # refactor chromosomes
    dplyr::mutate(
      chr = chr %>% {forcats::fct_reorder(., gsub('chr', '', .) %>% as.numeric())},
      direction = dplyr::case_when(
        state < 3 ~ 'loss',
        state > 3 ~ 'dupli',
        TRUE ~ 'neutral'),
      ploidy = dplyr::case_when(
        state == 1 ~ '0x',
        state == 2 ~ '0.5x',
        state == 3 ~ '1x',
        state == 4 ~ '1.5x',
        state == 5 ~ '2x',
        state == 6 ~ '>2x',
        TRUE ~ NA
        ) %>% forcats::fct_reorder(state),
      region = gsub('.*region_', '', cnv_name) %>% as.numeric(),
      cell_group = cell_group_name %>% gsub('\\..*', '', .))

@@

}, file = 'cnv_regions.rds', rerun = params$rerun)
```

```{r plot_cnv_regions_pred, fig.height = 20, fig.width = 10}
cnv_regions %>%
  ggplot(aes(x = midpoint, fill = ploidy, group = ploidy)) +
  geom_density(data = cnv_regions %>% dplyr::filter(direction == 'dupli'),
               aes(y = ..density..), 
               position = position_stack(), 
               colour = NA) +
  geom_density(data = cnv_regions %>% dplyr::filter(direction == 'loss'),
               aes(y = -..density..), 
               position = position_stack(reverse = T), 
               colour = NA) +
  ggh4x::facet_grid2(chr ~ cell_group, scales = 'free_x', independent = 'x') +
  theme_bw() +
  theme(axis.text.x = element_blank()) +
  scale_fill_manual(
    values = c('0x' = '#00008B', '0.5x' = '#b4b0ff', '1' = 'white', 
               '1.5x' = '#ffc2c2', '2x' = '#d64242', '>2x' = '#8B0000')) 
```

@@

```{r ct_comp, fig.width = 20, fig.height = 7}
p_dat <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::mutate(
    sample = forcats::fct_infreq(sample),
    cell_group = forcats::fct_infreq(cell_group)) %>%
  dplyr::add_count(sample, cell_group, name = 'n_sample_x_cell_group')
p_base <-
  p_dat %>%
  ggplot(aes(fill = cell_group, colour = cell_group)) 
p1 <- 
  p_base +
  geom_point(aes(x = sample, y = cell_group, 
                 alpha = n_sample_x_cell_group, size = n_sample_x_cell_group)) +
  theme_classic() +
  theme(legend.position = 'none') +
  scale_x_discrete(guide = guide_axis(angle = -90))
p2 <- 
  p_base +
  geom_bar(aes(x = sample), position = 'stack') +
  theme_void() +
  theme(legend.position = 'none') 
p3 <-
  p_base +
  geom_bar(aes(y = cell_group), position = 'stack') +
  theme_void() +
  theme(legend.position = 'none')
design <-
'BB#
AAC'
patchwork::wrap_plots(A = p1, B = p2, C = p3, design = design)
```

@@

```{r plot_umaps}
dittoSeq::dittoDimPlot(sce, params$cell_group_name, size = 0.2) +
  coord_fixed()
```

@@

  # plot both count (stack) and proportion (fill), annotate n cells
  list(
    plot_func(p_dat, position_fill(reverse = T)) +
      ggtitle(title) +
      theme(axis.text.x = element_blank(),
            axis.title.x = element_blank()) +
      ylab('count'),
    plot_func(p_dat, position_stack(reverse = T)) +
      ylab('proportion')) %>%
    patchwork::wrap_plots(ncol = 1) 
}

# plot, across all and by patient
plot_prop_chr7_gain(p_dat, 'chr7 gain - all cells')
p_dat %>%
  {split(., .$sample)} %>%
  purrr::walk2(names(.), ., function(id, df) {
    df %>%
    plot_prop_chr7_gain(
      paste0('chr7 gains - ', id, ' (', unique(df$n_cells), ' cells)')) %>%
      print()
  })
```

## Assigning whole chromosome-level events

@@

}, file = 'sce_w_gains.rds', rerun = params$rerun)
```

### Co-occurence of CNV events

@@

```{r chr7_and_others}
events <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::select(sample, cell, cell_group, 
                dplyr::ends_with('_gain'), 
                dplyr::ends_with('_loss')) %>%
  tidyr::pivot_longer(-c('sample', 'cell', 'cell_group')) %>%
  dplyr::group_by(sample, cell, cell_group) %>%
  # remove cells in which no calls made
  dplyr::filter(!all(is.na(value))) %>%
  dplyr::mutate(
    event = dplyr::case_when(
      all(value == 0 | is.na(value)) ~ 'WT',
      value == 1 ~ name,
      TRUE ~ NA)) %>%
  dplyr::summarise(
    events = paste(sort(unique(na.omit(event))), collapse = ','),
    n_events = sum(value, na.rm = T))

p_dat <-
  events %>%
  dplyr::ungroup() %>%
  dplyr::add_count(events) %>%
  dplyr::mutate(prop = n / dplyr::n()) %>%
  # ignore events that happen in <= 25 cells
  dplyr::filter(events != 'WT', n >= 25) %>%
  dplyr::mutate(
    events = events %>% 
      gsub('chr', '', .) %>% 
      gsub('_gain', '+', .) %>%
      gsub('_loss', '-', .),
    events = forcats::fct_infreq(events),
    label1 = scales::label_comma()(n), 
    label = dplyr::case_when(
      n < 10 ~ label1,
      TRUE ~ paste0(label1, ' (', round(prop*100, 2), '%)')
    ))

# plot
p <-
  p_dat %>%
  ggplot(aes(x = events)) +
  geom_text(aes(y = n + 50,  label = label), 
            vjust = 0.5, hjust = 0, angle = 90, size = 2) +
  scale_x_discrete(guide = guide_axis(angle = -90)) +
  ylim(c(0, max(p_dat$n) * 1.1)) +
  theme_classic() +
  theme(legend.position = 'bottom')
p + 
  geom_bar(aes(fill = sample))
p + 
  geom_bar(aes(fill = cell_group)) +
  theme(legend.position = 'none')
```

@@

         # return relevant stats
        tibble::tibble(
          OR = fisher$estimate, 
          p = fisher$p.value)

@@

        # no event
        tibble::tibble(OR = NA, p = NA)

      }

    }) %>%
    dplyr::bind_rows(.id = 'event') %>%
    # remove chr7+ (can't do chr7+ vs chr7+)
    dplyr::filter(event != 'chr7_gain') %>%
    dplyr::arrange(-OR)

@@

}, file = 'fisher.rds', rerun = params$rerun)

@@

@@

```{r chr7_gain_prop, results = 'asis'}
dat <-
  tibble::tibble(chr7_gain_value = sce$chr7_gain,
               cell_group = sce$cell_group) %>%
  dplyr::mutate(
    chr7_gain_status = dplyr::case_when(
      is.na(chr7_gain_value) ~ 'unknown/intermediate',
      chr7_gain_value == 0 ~ 'WT',
      chr7_gain_value == 1 ~ 'gain')
  )

cat('Counts and proportions of chr7 gain, all cells')
dat %>%
  dplyr::mutate(n_cells = dplyr::n()) %>%
  dplyr::group_by(chr7_gain_status, n_cells) %>%
  dplyr::summarise(n = dplyr::n(),
                   `%` = 100 * dplyr::n() / unique(n_cells)) %>%
  knitr::kable()

cat('Counts and proportions of chr7 gain, by cell group')
dat %>%
  dplyr::group_by(cell_group) %>%
  dplyr::mutate(n_cells = dplyr::n()) %>%
  dplyr::group_by(cell_group, chr7_gain_status, n_cells) %>%
  dplyr::summarise(n = dplyr::n(),
                   `%` = 100 * dplyr::n() / unique(n_cells)) %>%
  knitr::kable()
```

```{r pseudobulk_chr7_gain, class.source = 'fold-show'}
reduced_sce <- xfun::cache_rds({

@@

}, file = 'reduced_sce.rds', rerun = params$rerun)
```

@@

```{r deseq, message = T}
# fit one GLM model for each gene
fit <- xfun::cache_rds({
  fit <-
    glmGamPoi::glm_gp(
      reduced_sce,
      design = ~ chr7_gain + sample + cell_group,
      verbose = T
    )
  fit
}, file = 'fit.rds', rerun = params$rerun)

@@

  # define significant up/downregulation and position
  de_res <-
    de_res %>% 
    tibble::as_tibble() %>%
    dplyr::mutate(
      direction = dplyr::case_when(
        lfc < -1.5 & adj_pval < min_p ~ 'downregulated',
        lfc > 1.5 & adj_pval < min_p ~ 'upregulated',
        TRUE ~ NA)) %>%
    dplyr::left_join(coords)

@@

}, file = 'de_res.rds', rerun = params$rerun)
```

```{r plot_de}
de_res %>%
  ggplot(aes(x = lfc, y = -log10(adj_pval), colour = direction, 
             size = chr == 'chr7', shape = chr == 'chr7')) +
  geom_point() +
  geom_hline(yintercept = -log10(min_p), linetype = 'dashed') +
  geom_vline(xintercept = c(log2(1.5), log2(2)), linetype = 'dashed', colour = 'purple') +
  geom_vline(xintercept = c(-1.5, 1.5), linetype = 'dashed') +
  ggrepel::geom_text_repel(
    data = 
      de_res %>% 
      dplyr::filter(!is.na(direction)) %>%
      dplyr::group_by(direction) %>%
      dplyr::filter(rank(adj_pval) <= 10 | rank(lfc, ties.method = 'min') <= 10),
    aes(label = name, ),
    min.segment.length = 0, nudge_x = 0.7, show.legend = F) +
  theme_classic() +
  scale_colour_manual(values = c('upregulated' = 'darkred',
                                 'downregulated' = 'darkblue'), 
                      na.value = 'darkgrey') +
  scale_size_manual(values = c('TRUE' = 3, 'FALSE' = 2), na.value = 2) +
  scale_shape_manual(values = c('TRUE' = 55, 'FALSE' = 16), na.value = 16)
```

@@

seu <- xfun::cache_rds({
  seu <- Seurat::as.Seurat(sce)
  Seurat::Idents(seu) <- seu$cell_group
  seu <- Seurat::ScaleData(seu)
  seu
}, file = 'seu.rds', rerun = params$rerun)
```

@@

```{r markers}
markers <- xfun::cache_rds({
  markers <-
    Seurat::FindAllMarkers(seu)
  markers
}, file = 'markers.rds', rerun = params$rerun)
```

>>>>

---
title: "Gains and losses in rheumatoid arthritis data from inferCNV"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), "%d %B, %Y")`"
output:
  html_document:
    fig_width: 8
    keep_md: true
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
  sce: NULL
  cell_group_name: NULL
  infercnv_dir: NULL
  metadata: NULL
  cache_dir: NULL
  rerun: true
  max_prop_for_wt: 0.5
  min_prop_for_cnv: 0.8
---

```{r setup, include = F, message = F, warning = F, class.source = "fold-hide"}
# rmarkdown::render("reports/gains_and_losses.Rmd", output_file = "gains_and_losses.html", output_dir = "reports/", params = list(rerun = T))

@@

# chunk options
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(warning = FALSE, dpi = 300, message = FALSE,
                      cache.path = params$cache_dir)
```

@@

```{r load_sce, results = "asis", class.source = "fold-show"}
# load gene coords
coords <-
  "data/hg38_gencode_v27.txt" %>%
  readr::read_tsv(col_names = c("name", "chr", "start", "stop"))

# load CTAP and metadata
ctap <-
  "data/Zhang2023/syn26710600/CTAP_donor_mapping.tsv" %>%
  readr::read_tsv() %>%
  dplyr::select(sample = donor, CTAP)
md <-
  "data/Zhang2023/syn26710600/SYNAPSE_METADATA_MANIFEST.tsv" %>%
  readr::read_tsv() %>%
  dplyr::distinct()

@@

  # get cnv metadata
  cnvs <-
    list.files(params$infercnv_dir, pattern = "map_metadata_from_infercnv.txt",
               recursive = TRUE) %>%
    purrr::set_names(., dirname(.)) %>%
    purrr::map(function(file) {
      file.path(params$infercnv_dir, file) %>%
        read.table(sep = "\t")
    }) %>%
    dplyr::bind_rows() %>%
    dplyr::mutate(infercnv = TRUE)

  # left join cnvs to coldata, reorder, restore rownames
  col_data <- merge(sce@colData, cnvs, by = "row.names", all.x = TRUE,
                    sort = FALSE)
  rownames(col_data) <- col_data[, 1]
  col_data <-
    col_data[colnames(sce), -1] %>%
    DataFrame()

@@

}, file = "sce.rds", rerun = params$rerun)

@@

```{r echo = F, results = "asis"}
tibble::tribble(
  ~name, ~`type - value`,
  "`has_*`", "binary - whether any CNV / loss / duplication is found within the given chr",
  "`proportion_*`", "proportion - proportion of genes that are part of any CNV / loss / duplication within the given chr",
  "`proportion_scaled_*`", "proportion - same as proportion_*, but taking into account whether genes that are duplicated / lost are a single copy or two copies, as a weight",
  "`top_loss_*`", "the top CNVs that affect the most genes"
) %>%
  knitr::kable()
```

@@

  # get HMM predictions
  cnv_regions <-
    list.files(
      params$infercnv_dir,
      pattern = "HMM_CNV_predictions.HMMi6.leiden.hmm_mode-subclusters.Pnorm_0.5.pred_cnv_regions.dat",
      recursive = T) %>%
    purrr::set_names(., dirname(.)) %>%
    purrr::map(function(file) {
      file.path(params$infercnv_dir, file) %>% 
        readr::read_tsv(show_col_types = F)
      }) %>%
    dplyr::bind_rows(.id = "sample") %>%
    # refactor chromosomes
    dplyr::mutate(
      chr = chr %>% {forcats::fct_reorder(., gsub("chr", "", .) %>% as.numeric())},
      direction = dplyr::case_when(
        state < 3 ~ "loss",
        state > 3 ~ "dupli",
        TRUE ~ "neutral"),
      ploidy = dplyr::case_when(
        state == 1 ~ "0x",
        state == 2 ~ "0.5x",
        state == 3 ~ "1x",
        state == 4 ~ "1.5x",
        state == 5 ~ "2x",
        state == 6 ~ ">2x",
        TRUE ~ NA
        ) %>% forcats::fct_reorder(state),
      region = gsub(".*region_", "", cnv_name) %>% as.numeric(),
      cell_group = cell_group_name %>% gsub("\\..*", "", .))

@@

}, file = "cnv_regions.rds", rerun = params$rerun)
```

```{r plot_cnv_regions_pred, fig.height = 20, fig.width = 10}
cnv_regions %>%
  ggplot(aes(x = midpoint, fill = ploidy, group = ploidy)) +
  geom_density(data = cnv_regions %>% dplyr::filter(direction == "dupli"),
               aes(y = ..density..), 
               position = position_stack(), 
               colour = NA) +
  geom_density(data = cnv_regions %>% dplyr::filter(direction == "loss"),
               aes(y = -..density..), 
               position = position_stack(reverse = T), 
               colour = NA) +
  ggh4x::facet_grid2(chr ~ cell_group, scales = "free_x", independent = "x") +
  theme_bw() +
  theme(axis.text.x = element_blank()) +
  scale_fill_manual(
    values = c("0x" = "#00008B", "0.5x" = "#b4b0ff", "1" = "white", 
               "1.5x" = "#ffc2c2", "2x" = "#d64242", ">2x" = "#8B0000")) 
```

@@

```{r ct_comp, fig.width = 12, fig.height = 7}
p_dat <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::mutate(
    sample = forcats::fct_infreq(sample),
    cell_group = forcats::fct_rev(forcats::fct_infreq(cell_group))) %>%
  dplyr::add_count(sample, cell_group, name = "n_sample_x_cell_group") %>%
  dplyr::add_count(cell_group, name = "n_cell_group")
p_base <-
  p_dat %>%
  ggplot(aes(fill = cell_group, colour = cell_group)) +
  scale_fill_manual(values = dittoSeq::dittoColors()) +
  scale_colour_manual(values = dittoSeq::dittoColors())
p1 <- 
  p_base +
  geom_point(aes(x = sample, y = cell_group, 
                 alpha = n_sample_x_cell_group, size = n_sample_x_cell_group)) +
  theme_classic() +
  theme(legend.position = "none",
        axis.text.x = element_text(size = 7)) +
  scale_x_discrete(guide = guide_axis(angle = -90)) +
  labs(y = params$cell_group_name)
p2 <- 
  p_base +
  geom_bar(aes(x = sample), position = "stack") +
  theme_void() +
  theme(legend.position = "none",
        axis.text.y = element_text()) 
p3 <-
  p_base +
  geom_bar(aes(y = cell_group), position = "stack") +
  geom_text(
    data = p_dat %>% dplyr::distinct(cell_group, n_cell_group),
    aes(x = n_cell_group, y = cell_group, 
        label = format(n_cell_group, nsmall = 1, big.mark = ",")), 
    hjust = -0.1) +
  theme_void() +
  theme(legend.position = "none") +
  xlim(0, 1.2 * max(p_dat$n_cell_group))
design <-
"BB#
BB#
AAC"

# plot
patchwork::wrap_plots(A = p1, B = p2, C = p3, design = design)
```

@@

```{r plot_umaps}
dittoSeq::dittoDimPlot(sce, params$cell_group_name, size = 0.2) +
  coord_fixed()

pal <-
  c("T cell" = "#E6AB02",
    "Stromal cell" = "#66A61E",
    "Myeloid cell" = "#E7298A",
    "B cell/plasma cell" = "#7570B3",
    "Endothelial cell" = "#D95F02",
    "NK" = "#1B9E77")
dittoSeq::dittoDimPlot(sce, "cell_type", size = 0.2) +
  coord_fixed() +
  scale_colour_manual(values = pal)
```

@@

  # plot both count (stack) and proportion (fill), annotate n cells
  list(
    plot_func(p_dat, position_fill(reverse = T)) +
      ggtitle(title) +
      theme(axis.text.x = element_blank(),
            axis.title.x = element_blank()) +
      ylab("proportion"),
    plot_func(p_dat, position_stack(reverse = T)) +
      ylab("count")) %>%
    patchwork::wrap_plots(ncol = 1) 
}

# plot, across all and by patient
plot_prop_chr7_gain(p_dat, "chr7 gain - all cells")
p_dat %>%
  {split(., .$sample)} %>%
  purrr::walk2(names(.), ., function(id, df) {
    df %>%
    plot_prop_chr7_gain(
      paste0("chr7 gains - ", id, " (", unique(df$n_cells), " cells)")) %>%
      print()
  })
```

# Assigning whole chromosome-level events

@@

}, file = "sce_w_gains.rds", rerun = params$rerun)
```

```{r plot_heatmap, fig.width = 10, fig.height = 5}
event_cols <-
  c("WT" = "grey",
    "chr5_gain" = "#e7298a",
    "chr5_gain,chr7_gain" = "#d95f02",
    "chr7_gain" = "#e6ab02")

p_dat <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::select(sample, cell, 
                dplyr::ends_with("_gain"), 
                dplyr::ends_with("_loss")) %>%
  tidyr::pivot_longer(-c("sample", "cell")) %>%
  dplyr::filter(name %in% c("chr7_gain", "chr5_gain")) %>%
  dplyr::group_by(sample, cell) %>%
  # remove cells in which no calls made
  dplyr::filter(!all(is.na(value))) %>%
  dplyr::mutate(
    event = dplyr::case_when(
      all(value == 0 | is.na(value)) ~ "WT",
      value == 1 ~ name,
      TRUE ~ NA)) %>%
  dplyr::summarise(
    events = paste(sort(unique(na.omit(event))), collapse = ",")) %>%
  dplyr::mutate(events = forcats::fct_relevel(events, names(event_cols))) %>%
  # get n aneuploid for plot ordering
  dplyr::group_by(sample) %>%
  dplyr::mutate(prop_aneuploid = sum(events != "WT") / dplyr::n()) %>%
  # count
  dplyr::group_by(sample, prop_aneuploid) %>%
  dplyr::count(events) %>%
  # for fill and stack
  tidyr::crossing(tibble::tibble(plot = c("% cells", "n cells")))

# plot
p_dat %>%
  ggplot(aes(x = reorder(sample, -prop_aneuploid), y = n, fill = events)) +
  geom_col(data = p_dat %>% dplyr::filter(plot == "% cells"),
           position = "fill") +
  geom_col(data = p_dat %>% dplyr::filter(plot == "n cells"),
           position = "stack") +
  theme_classic() +
  ggh4x::facet_grid2(plot ~ ., scales = "free_y", independent = "y") +
  scale_x_discrete(guide = guide_axis(angle = -90)) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_manual(values = event_cols) +
  theme(axis.title = element_blank(),
        axis.text.x = element_text(size = 6))
```

## Co-occurence of CNV events

@@

```{r chr7_and_others}
events <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::select(sample, cell, cell_group, 
                dplyr::ends_with("_gain"), 
                dplyr::ends_with("_loss")) %>%
  tidyr::pivot_longer(-c("sample", "cell", "cell_group")) %>%
  dplyr::group_by(sample, cell, cell_group) %>%
  # remove cells in which no calls made
  dplyr::filter(!all(is.na(value))) %>%
  dplyr::mutate(
    event = dplyr::case_when(
      all(value == 0 | is.na(value)) ~ "WT",
      value == 1 ~ name,
      TRUE ~ NA)) %>%
  dplyr::summarise(
    events = paste(sort(unique(na.omit(event))), collapse = ","),
    n_events = sum(value, na.rm = T))

p_dat <-
  events %>%
  dplyr::ungroup() %>%
  dplyr::add_count(events) %>%
  dplyr::mutate(prop = n / dplyr::n()) %>%
  # ignore events that happen in <= 25 cells
  dplyr::filter(events != "WT", n >= 25) %>%
  dplyr::mutate(
    events = events %>% 
      gsub("chr", "", .) %>% 
      gsub("_gain", "+", .) %>%
      gsub("_loss", "-", .),
    events = forcats::fct_infreq(events),
    label1 = scales::label_comma()(n), 
    label = dplyr::case_when(
      n < 10 ~ label1,
      TRUE ~ paste0(label1, " (", round(prop*100, 2), "%)")
    ))

# plot
p <-
  p_dat %>%
  ggplot(aes(x = events)) +
  geom_text(aes(y = n,  label = label), 
            vjust = 0.5, hjust = 0, angle = 90, size = 2) +
  scale_x_discrete(guide = guide_axis(angle = -90)) +
  ylim(c(0, max(p_dat$n) * 1.1)) +
  theme_classic() +
  theme(legend.position = "bottom")
p + 
  geom_bar(aes(fill = sample))
p + 
  geom_bar(aes(fill = cell_group)) +
  theme(legend.position = "none")
```

### Fisher exact test

@@

        # get contingencies
        cont <-
          table(sce@colData[, c("chr7_gain", event)])

         # return relevant stats
        tibble::tibble(
          OR = fisher$estimate, 
          p = fisher$p.value,
          sig = ifelse(fisher$p.value < 0.05, "*", ""),
          n_WT = cont[1, 1],
          n_chr7_gain = cont[2, 1],
          n_event = cont[1, 2],
          n_both = cont[2, 2])

@@

        # no event
        tibble::tibble()
      }

    }) %>%
    dplyr::bind_rows(.id = "event") %>%
    # remove chr7+ (can't do chr7+ vs chr7+)
    dplyr::filter(event != "chr7_gain", n_event > 0) %>%
    dplyr::arrange(p)

@@

}, file = "fisher.rds", rerun = T)

@@

### GLM

We can also run a GLM.

```{r glm}
# read in metadata
metadata <-
  readr::read_tsv(params$metadata)

# create glm
# prop_chr7_gain ~ age (+ coverage) (per donor)
for_glm <-
  dplyr::inner_join(
    tibble::as_tibble(sce@colData), 
    metadata, 
    by = c("sample" = "id")) %>%
  dplyr::filter(!is.na(age), !is.na(chr7_gain)) %>%
  dplyr::group_by(sample, age, CTAP, sex, race, ethnicity) %>%
  dplyr::summarise(n_chr7_gain = sum(chr7_gain),
                   n_cells = dplyr::n(),
                   ch7_gain_prop = n_chr7_gain / n_cells)

for_glm %>%
  {glm(.$n_chr7_gain ~ log(.$n_cells) + .$age)}
```

```{r plot_glm_vars}
# n cells vs n chr7 gain
for_glm %>%
  ggplot(aes(x = n_cells, y = n_chr7_gain)) +
  geom_point() +
  theme_classic()

# age vs prop chr7 gain
for_glm %>%
  ggplot(aes(x = age, y = ch7_gain_prop)) +
  geom_point() +
  theme_classic()

# all categorical variables
for_glm %>%
  tidyr::pivot_longer(cols = c("CTAP", "sex", "race", "ethnicity")) %>%
  dplyr::group_by(name, value) %>%
  dplyr::mutate(med = median(ch7_gain_prop)) %>%
  ggplot(aes(
    x = tidytext::reorder_within(value, by = med, within = name), 
    y = ch7_gain_prop)) +
  geom_boxplot() +
  geom_jitter(height = 0) +
  facet_wrap(~ name, nrow = 1, scales = "free") +
  tidytext::scale_x_reordered(guide = guide_axis(angle = -45)) +
  theme_classic() +
  labs(x = "")

# looking by cluster
for_glm_w_cluster <-
  dplyr::inner_join(
    tibble::as_tibble(sce@colData), 
    metadata, 
    by = c("sample" = "id")) %>%
  dplyr::filter(!is.na(age), !is.na(chr7_gain)) %>%
  dplyr::group_by(
    sample, age, CTAP, sex, race, ethnicity, cluster_name) %>%
  dplyr::summarise(n_chr7_gain = sum(chr7_gain),
                   n_cells = dplyr::n(),
                   ch7_gain_prop = n_chr7_gain / n_cells) 
for_glm_w_cluster %>%
  ggplot(aes(x = cluster_name, y = ch7_gain_prop)) +
  geom_violin() +
  geom_jitter(height = 0) +
  theme_classic() +
  scale_x_discrete(guide = guide_axis(angle = -45))
```

@@

```{r chr7_gain_prop, results = "asis"}
dat <-
  tibble::tibble(chr7_gain_value = sce$chr7_gain,
               cell_group = sce$cell_group) %>%
  dplyr::mutate(
    chr7_gain_status = dplyr::case_when(
      is.na(chr7_gain_value) ~ "unknown/intermediate",
      chr7_gain_value == 0 ~ "WT",
      chr7_gain_value == 1 ~ "gain")
  )

cat("Counts and proportions of chr7 gain, all cells")
dat %>%
  dplyr::mutate(n_cells = dplyr::n()) %>%
  dplyr::group_by(chr7_gain_status, n_cells) %>%
  dplyr::summarise(n = dplyr::n(),
                   `%` = 100 * dplyr::n() / unique(n_cells)) %>%
  knitr::kable()

cat("Counts and proportions of chr7 gain, by cell group")
dat %>%
  dplyr::group_by(cell_group) %>%
  dplyr::mutate(n_cells = dplyr::n()) %>%
  dplyr::group_by(cell_group, chr7_gain_status, n_cells) %>%
  dplyr::summarise(n = dplyr::n(),
                   `%` = 100 * dplyr::n() / unique(n_cells)) %>%
  knitr::kable()
```

```{r pseudobulk_chr7_gain, class.source = "fold-show"}
reduced_sce <- xfun::cache_rds({

@@

}, file = "reduced_sce.rds", rerun = params$rerun)
```

@@

```{r deseq, message = T}
# fit one GLM model for each gene
fit <- xfun::cache_rds({
  fit <-
    glmGamPoi::glm_gp(
      reduced_sce,
      design = ~ chr7_gain + sample + cell_group,
      verbose = T
    )
  fit
}, file = "fit.rds", rerun = params$rerun)

@@

  # define significant up/downregulation and position
  de_res <-
    de_res %>% 
    tibble::as_tibble() %>%
    dplyr::mutate(
      direction = dplyr::case_when(
        lfc < -1.5 & adj_pval < min_p ~ "downregulated",
        lfc > 1.5 & adj_pval < min_p ~ "upregulated",
        TRUE ~ NA)) %>%
    dplyr::left_join(coords)

@@

}, file = "de_res.rds", rerun = params$rerun)
```

```{r plot_de}
de_res %>%
  ggplot(aes(x = lfc, y = -log10(adj_pval), colour = direction, 
             size = chr == "chr7", shape = chr == "chr7")) +
  geom_point() +
  geom_hline(yintercept = -log10(min_p), linetype = "dashed") +
  geom_vline(xintercept = c(log2(1.5), log2(2)), linetype = "dashed", colour = "purple") +
  geom_vline(xintercept = c(-1.5, 1.5), linetype = "dashed") +
  ggrepel::geom_text_repel(
    data = 
      de_res %>% 
      dplyr::filter(!is.na(direction)) %>%
      dplyr::group_by(direction) %>%
      dplyr::filter(rank(adj_pval) <= 10 | rank(lfc, ties.method = "min") <= 10),
    aes(label = name),
    min.segment.length = 0, nudge_x = 0.7, show.legend = F) +
  theme_classic() +
  scale_colour_manual(values = c("upregulated" = "darkred",
                                 "downregulated" = "darkblue"), 
                      na.value = "darkgrey") +
  scale_size_manual(values = c("TRUE" = 3, "FALSE" = 2), na.value = 2) +
  scale_shape_manual(values = c("TRUE" = 55, "FALSE" = 16), na.value = 16)
```

```{r plot_de_close}
min_p_non_chr7 <-
  de_res %>%
  dplyr::filter(!is.na(direction)) %>%
  dplyr::pull(adj_pval) %>% 
  min()
de_res %>%
  ggplot(aes(x = lfc, y = -log10(adj_pval), colour = direction, 
             size = chr == "chr7", shape = chr == "chr7")) +
  geom_point() +
  geom_hline(yintercept = -log10(min_p), linetype = "dashed") +
  geom_vline(xintercept = c(log2(1.5), log2(2)), linetype = "dashed", colour = "purple") +
  geom_vline(xintercept = c(-1.5, 1.5), linetype = "dashed") +
  ggrepel::geom_text_repel(
    data = 
      de_res %>% 
      dplyr::filter(!is.na(direction)) %>%
      dplyr::group_by(direction) %>%
      dplyr::filter(rank(adj_pval) <= 10 | rank(lfc, ties.method = "min") <= 10),
    aes(label = name),
    min.segment.length = 0, nudge_x = 0.7, show.legend = F) +
  theme_classic() +
  scale_colour_manual(values = c("upregulated" = "darkred",
                                 "downregulated" = "darkblue"), 
                      na.value = "darkgrey") +
  scale_size_manual(values = c("TRUE" = 5, "FALSE" = 4), na.value = 4) +
  scale_shape_manual(values = c("TRUE" = 55, "FALSE" = 16), na.value = 16) +
  ylim(0, -log10(min_p_non_chr7) * 1.03)
```

@@

seu <- xfun::cache_rds({
  seu <- Seurat::as.Seurat(sce)
  Seurat::Idents(seu) <- seu$cell_group
  seu <- Seurat::ScaleData(seu)
  seu
}, file = "seu.rds", rerun = params$rerun)
```

@@

```{r markers}
markers <- xfun::cache_rds({
  markers <-
    Seurat::FindAllMarkers(seu)
  markers
}, file = "markers.rds", rerun = params$rerun)
```
